Topological Equivariant Artist Model for Visualization Library Architecture

### 2024-01-12 paper/main.tex
equivariance figure

\IEEEPARstart{V}{isualization} tools implement transforms of the form $data \rightarrow graphic$ and are expected to be structure preserving. Loosely, that means the mathematical properties of the data and how the points are connected to each other should be inferable such that  $graphic \rightarrow data$. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Motivated by wanting to make better tools, we use algebraic topology to describe how $data$ and a corresponding $graphics$ have the same underlying structure and we use category theory to describing under what conditions the visualization tool preserves that structure such that $graphic \leftrightarrow data$.

Algebraic structures allow us to describe many types of inherent structure in a uniform manner, and category theory provides a rigorous framework for specifying the behavior and properties of structure and maps between structures. For example, our framework can encapsulate how a table and scatter plot and heatmap are different representations of the same data and track an observation from a datacube as a point along a time series and on a map and in a network. The algebraic structures can be translated into programmatic types, while the categorical descriptions translate to a functional design framework. Strong typing and function composition enable library developers to build complex components from simpler verifiable parts \cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These  components can be built as a standalone library and integrated into existing libraries and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib.

@@

Since the earliest maps \cite{friendlyBriefHistoryData2006}, a visualization has been expected to preserve the $field$ properties and $topology$ of the corresponding dataset; a \textcolor{fiber}{\textbf{field}} \footnote{Throughout this paper, we augment the math with color to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}. This color coding carries through to the figures.}  is a set of values of the same type, e.g. one column of a table or the pixels of an image, and \textcolor{base}{\textbf{topology}} is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}. In \autoref{fig:related-work:continuity:ktypes}, the topology is a continuous 3 dimensional surface encapsulating time and position, while the fields are date, location, station, pressure, and temperature. Our paper provides a language to express how the scatter plot, line plot, and heatmap are preserving different structural properties of the dataset, thereby generalizing Bertin's codification of how the topology of observations must match the class of representation (i.e. point , line, area) and how the variation possible in a field match the variation in the corresponding visual variable \cite{bertinSemiologyGraphicsDiagrams2011} and Mackinlay's work on how structure preservation includes preserving the mathimatical relations between elements of the same and different fields \cite{mackinlayAutomatingDesignGraphical1986}.

While Bertin and Mackinlay's work primarily discusses the $data \rightarrow graphic$, Kindlemann and Scheidegger's algebraic framework \cite{kindlmannAlgebraicProcessVisualization2014} also dictates a set of conditions under which the $graphic \rightarrow data$  mapping is structure preserving. Encapsulating their conditions, we: present a uniform abstract data representation layer in \autoref{sec:atct:sheaves} for ensuring that the visualization should not change if the data representation (i.e. the data container) changes, define the conditions under which data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009} in \autoref{sec:atct:xi}, and provide a methodology for verifying that changes in data should correspond to changes in the visualization in \autoref{sec:artist:equivariant:artist} that does not necessarily require that the changes be perceptually significant. Furthermore, our model extends their conditions of equivariance by explicitly incorporating topology and by providing a framework for translating the theoretical ideas into buildable components in \autoref{sec:construction}.

@@

where the measurements are the groups and changes to them are actions. Structure preservetion in fields means that the data and the visualization are equivalent representations of an underlying phenomena such that changes to data or visualization be reflected in the other in a consistent way. A function that preserves structure in this manner for groups is called equivariant. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$
\begin{definition}\label{def:equivariance}
 A function $f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
\end{definition}
which means that a visualization is structure preserving when there exist compatible group actions on the data and visualization, as proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

Some fields have structures that cannot be codified as groups, for example the ordinal 
scale is a type of structure called a monoid\footnote{The difference between a monoid 
and a group is that monoid operators do not have to be invertable.} For these situations, 
we build on the claim that visualizations are $homomorphic$, as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. Given the function $f: X \rightarrow Y$, with
compatiable operators $(X,*)$ \note{finish this out}

\begin{definition}\label{def:homomorphism}
  A function $f$ is \textbf{homomorphic} when

  $f(act_x(x)) = act_y(f(x))$ for all $x, y \in X$ if there exists a compatibility function $\varphi(act_y) = act_x$.

\end{definition}
which means that applying the operation $*$ to the input to $f$ is equivalent to applying the operation $\circ$ to the output.

@@

\section{Conclusion}
The toy example presented in \autoref{sec:discussion} demonstrates that it is relatively straightforward to build working visualization library components using the construction described in \autoref{sec:construction}. Since these components are defined with single record inputs, the can be implemented such that they are concurrent. The cost of building a new function using these components is sometimes as small as renaming fields, meaning the new feature is relatively easy to maintain. These new components are also a lower maintenance burden because, by definition, they are designed in conjunction with tests that verify that they are equivariant.  
These new components are also compatible with the existing library architecture, allowing for a slow iterative transition to components built using this framework. 
The framework introduced in this paper is a marriage of the ways the graphic and data visualization communities approach visualization. The graphic community prioritizes \note{?} how input is translated to output, which is encapsulated in the artist $\vartist$. The data visualization community prioritizes the manner in which that input is encoded, which is encapsulated in the separation of stages $\vmark \circ \vchannel$. Formalizing that both views are equivalent $\vartist=\vmark \circ \vchannel$ gives library developers the flexibility to build visualization components in the manner that makes more sense for the domain without having to sacrifice the equivariance of the translation.

>>>>

\IEEEPARstart{V}{isualization} tools implement transforms of the form $data \rightarrow graphic$ and are expected to be structure preserving. Loosely, that means the mathematical properties of the data and how the points are connected to each other should be inferable such that  $graphic \rightarrow data$. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Therefore a visualization tool is structure preserving when it preserves the mapping $data\leftrightarrow graphic$.

We propose that we can better enforce this expectation in software by providing a uniform way of expressing $data$ and $graphic$ using their respective algebraic structure and by uniformally specifying the behaviors and properties of those structures and the maps between them using category theory. For example, our framework can encapsulate how a table and scatter plot and heatmap are different representations of the same data and track an observation from a datacube as a point along a time series and on a map and in a network. The algebraic structures can then be translated into programmatic types, while the categorical descriptions translate to a functional design framework. Strong typing and function composition enable visualization software developers to build complex components from simpler verifiable parts \cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These  components can be built as a standalone library and integrated into existing libraries and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib.

@@

Since the earliest maps \cite{friendlyBriefHistoryData2006}, a visualization has been expected to preserve the $field$ properties and $topology$ of the corresponding dataset:
\begin{description}
  \item[\textcolor{fiber}{\textbf{field}} \footnote{Throughout this paper, we augment the math with color to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}. This color coding carries through to the figures.}]  is a set of values of the same type, e.g. one column of a table or the pixels of an image
  \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}. 
\end{description}

 In \autoref{fig:related-work:continuity:ktypes}, the topology is a continuous 3 dimensional surface encapsulating time and position, while the fields are date, location, station, pressure, and temperature. Our paper provides a language to express how the scatter plot, line plot, and heatmap are preserving different structural properties of the dataset, thereby generalizing Bertin's codification of how the topology of observations must match the class of representation (i.e. point , line, area) and how the variation possible in a field match the variation in the corresponding visual variable \cite{bertinSemiologyGraphicsDiagrams2011} and Mackinlay's work on how structure preservation includes preserving the mathimatical relations between elements of the same and different fields \cite{mackinlayAutomatingDesignGraphical1986}.

While Bertin and Mackinlay's work primarily discusses the $data \rightarrow graphic$, Kindlemann and Scheidegger's algebraic framework \cite{kindlmannAlgebraicProcessVisualization2014} also dictates a set of conditions under which the $graphic \rightarrow data$  mapping is structure preserving. Encapsulating their conditions, we: present a uniform abstract data representation layer in \autoref{sec:atct:sheaves} for ensuring that the visualization should not change if the data representation (i.e. the data container) changes, define the conditions under which data is mapped unambiguously to visual encodings \cite{ziemkiewiczEmbeddingInformationVisualization2009} in \autoref{sec:atct:xi}, and provide a methodology for verifying that changes in data should correspond to changes in the visualization in \autoref{sec:artist:equivariant:artist} that does not necessarily require that the changes be perceptually significant. Furthermore, our model extends their conditions of equivariance by explicitly incorporating topology and by providing a framework for translating the theoretical ideas into buildable components in \autoref{sec:construction}.

@@

where the measurements are the groups and changes to them are actions. A function that preserves structure when the input or output is changed by a group action is called *equivariant*. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$
\begin{definition}\label{def:equivariance}
 A function $f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
\end{definition}
which means that a visualization is structure preserving when there exist compatible group actions on the data and visualization, as proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

While *equivariance* allows us to test stucture preservation under transformations to data or visualization, *homomorphism* provides a way of asserting whether the relations between data elements was preserved as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. Given the function $f: X \rightarrow Y$, with operators $(X, \circ)$ and $(Y, *)$

\begin{definition}\label{def:homomorphism}
  A function $f$ is \textbf{homomorphic} when $f(x_1 \circ x_2) = f(x_1) * f(x_2)$ and preserves identites $f(I_x) = I_y$ all $x, y \in X$

\end{definition}
which means that the operators $\circ$ and $*$ are compatiable. For example, although the Steven's scales were conceptualized as having group structure, the ordinal scale have a monoidal structure because partial orders ($\geq, \leq$) are not invertable. \note{check grimalidi for citation}.

As illustrated in \ref{fig:def:homomorphism}, .... structure preservation under ordering means that since 1 is less than 2, the visual transformation of 1 must be less than the visual transformation of two - here shown as 1 having a shorter bar than 2. We must define a < for visual compatable with the < for numbers. We handwave that for every relation we can write a compatabile closed function,  e.g. we can replace $\leq$ with $\min$, such that the operation is closed.

@@

\section{Conclusion}
The toy example presented in \autoref{sec:discussion} demonstrates that it is relatively straightforward to build working visualization library components using the construction described in \autoref{sec:construction}. Since these components are defined with single record inputs, they can be implemented such that they are concurrent. The cost of building a new function using these components is sometimes as small as renaming fields, meaning the new feature is relatively easy to maintain. These new components are also a lower maintenance burden because, by definition, they are designed in conjunction with tests that verify that they are equivariant.  
These new components are also compatible with the existing library architecture, allowing for a slow iterative transition to components built using this framework. 
The framework introduced in this paper is a marriage of the ways the graphic and data visualization communities approach visualization. The graphic community prioritizes \note{?} how input is translated to output, which is encapsulated in the artist $\vartist$. The data visualization community prioritizes the manner in which that input is encoded, which is encapsulated in the separation of stages $\vmark \circ \vchannel$. Formalizing that both views are equivalent $\vartist=\vmark \circ \vchannel$ gives library developers the flexibility to build visualization components in the manner that makes more sense for the domain without having to sacrifice the equivariance of the translation.
